### 2021-06-18 papers/thesis/full/ch2-spherical-harmonics/spherical-harmonics.tex
minor edits after readthrough (typos, repetition of accronym defs, etc.)

Before we proceed further, let us build up to our definition of the spherical harmonics. We first need to introduce a few classical orthogonal polynomials. \enquote{Orthogonal polynomials} (OPs) are defined as such by the condition that each polynomial is orthogonal to each other, with respect to a given inner product. In particular, they are orthogonal with respect to all lower degree polynomials.

@@

An important property of the Jacobi matrices is that they are sparse -- specifically they are \textit{banded-block-banded matrices}.

@@

Three-term recurrence relations for orthogonal polynomials are well established (see for example \cite[Section 18.9]{DLMF}), including for multidimensional OPs \cite[Section 3.2]{dunkl2014orthogonal}. In a similar vein, we can obtain a recurrence relation for the spherical harmonics by combining each system in \bsrefeqn{eqn:jacobimatsdef} in a specific way. It should be emphasised at this point that this is not necessarily the optimal way of computing the spherical harmonics; various efficient methods and algorithms have been proposed (see e.g. \cite{guy1998efficient, sloan2013efficient, fukushima2012numerical}). We present this method in order to be able to generalise it to other multidimensional OP families later on in this thesis.

@@

For each $l = 0,1,2\dots$ let $\Dlt$ be any matrix that is a left inverse of $A_l$, i.e. such that $\Dlt A_l = \idmat{2l+3}$. Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the $\Dlt$'s, we obtain a lower triangular system \cite[Section 3.2]{dunkl2014orthogonal}, which can be expanded to obtain the three-term recurrence, for $l = 0,1,2,\dots$:
\bseqn{
	\begin{cases}
		\SHvec_{-1}(x,y,z) := 0 \\
		\SHvec_{0}(x,y,z) := Y_0 \\
		\SHvec_{l+1}(x,y,z) = -\Dlt (B_l-G_l(x,y,z)) \SHvec_l(x,y,z) - \Dlt C_l  \,\SHvec_{l-1}(x,y,z)
	\end{cases}
}

@@

For the purposes of this discussion, we will only consider the tangent bundle, and thus wish to write down a basis for vector-valued functions with values in the tangent bundle. Fortunately, there is a simple solution known as the \textit{vector spherical harmonics (VSHs)} \cite{barrera1985vector}: \nomenclature[grad]{$\gradS$}{The spherical gradient operator} \nomenclature[Psilm]{$\gradYlm$}{The vector spherical harmonic $\gradS \Ylm$} \nomenclature[Philm]{$\gradpYlm$}{The vector spherical harmonic $\gradS^\perp \Ylm = \rvec \times \gradS \Ylm$}

@@

On that note, we will first consider the case that $m > 0$ and $m' = m - 1$, and calculate the integral in \bsrefeqn{eqn:lemmaVSHrecurrencesproof}:
\bseqnnumber{
	&\int_{-1}^{1} \Big\{ m(m-1) \jaclm{l}{m} \jaclmbracket{l'}{m-1} \rho^{2m-2} (1 + z^2) - m z \jaclm{l}{m} \jaclmbracketdz{l'}{m-1} \rho^{2m} \nonumber \\
	&\quadtwo \quad - (m-1) z \jaclmdz{l}{m} \jaclmbracket{l'}{m-1} \rho^{2m} + \jaclmdz{l}{m} \jaclmbracketdz{l'}{m-1} \rho^{2m+2} \Big\} \: \D z \nonumber \\
	&= \int_{-1}^{1} \Big\{ m(m-1) \jaclm{l}{m} \jaclmbracket{l'}{m-1} \rho^{2m-2} (1 + z^2) - m z \jaclm{l}{m} \jaclmbracketdz{l'}{m-1} \rho^{2m} \nonumber \\
	&\quadtwo + (m-1) \jaclm{l}{m} \rho^{2m-2} \big(\jaclmbracket{l'}{m-1} \rho^{2} - z \jaclmbracketdz{l'}{m-1} \rho^{2} - 2m z^2 \jaclmbracket{l'}{m-1} \big) \nonumber \\
	&\quadtwo + \jaclmdz{l}{m} \jaclmbracketdz{l'}{m-1} \rho^{2m+2} \Big\} \: \D z \nonumber \\
	&= \int_{-1}^{1} \Big\{ (m-1)(m+1) \jaclm{l}{m} \jaclmbracket{l'}{m-1} \rho^{2m} - z \jaclm{l}{m} \jaclmbracketdz{l'}{m-1} \rho^{2m} \nonumber \\
	&\quadfour + \jaclmdz{l}{m} \jaclmbracketdz{l'}{m-1} \rho^{2m+2} \Big\} \: \D z \nonumber \\
	&= \int_{-1}^{1} \Big\{ (m-1)(m+1) \jaclm{l}{m} \jaclmbracket{l'}{m-1} \rho^{2m} - d_{l',m-1} z \jaclm{l}{m} \jaclm{l'}{m} \rho^{2m} \nonumber \\
	&\quadfour + d_{l,m} d_{l',m-1} \jaclmbracket{l}{m+1} \jaclm{l'}{m} \rho^{2m+2} \Big\} \: \D z \nonumber \\
	&= \int_{-1}^{1} \Big\{ (m^2 - 1) \jaclm{l}{m} \big( \tilde \alpha_{l', m-1, 2} \jaclm{l'-1}{m} + \tilde \alpha_{l', m-1, 4} \jaclm{l'+1}{m} \big) \rho^{2m} \nonumber \\
	&\quadfour - d_{l',m-1} \jaclm{l}{m} \big( \tilde \gamma_{l',m,1} \jaclm{l'-1}{m} + \tilde \gamma_{l',m,2} \jaclm{l'+1}{m} \big) \rho^{2m} \nonumber \\
	&\quadfour + d_{l,m} d_{l',m-1} \jaclm{l'}{m} \big( \tilde \alpha_{l, m+1, 1} \jaclm{l-1}{m} + \tilde \alpha_{l, m+1, 3} \jaclm{l+1}{m} \big) \rho^{2m} \Big\} \: \D z \nonumber \\
	&= {\delta_{l',l-1} \over 2\pi} \Big\{ \big[ (m^2 - 1) \tilde \alpha_{l-1, m-1, 4} - d_{l-1,m-1} \tilde \gamma_{l-1,m,2} \big] {1 \over (\clm)^2} + d_{l,m} d_{l-1,m-1} {\tilde \alpha_{l-1, m-1, 1} \over (c_{l-1}^m)^2} \Big\} \nonumber \\
	&\quadtwo + {\delta_{l',l+1} \over 2\pi} \Big\{ \big[ (m^2 - 1) \tilde \alpha_{l+1, m-1, 2} - d_{l+1,m-1} \tilde \gamma_{l+1,m,1} \big] {1 \over (\clm)^2} \nonumber \\
	&\quadeight + d_{l,m} d_{l+1,m-1} {\tilde \alpha_{l-1, m-1, 3} \over (c_{l+1}^m)^2} \Big\}. \label{eqn:lemmaVSHrecurrencesproof2}
}
As we can see, the coefficients are then also only possibly nonzero for $l' = l \pm 1$. Combining the above with the \bsrefeqn{eqn:lemmaVSHrecurrencesproof}, we have that our two coefficients for $\calA_{l-1,m-1}, \calA_{l+1,m-1}$ are those stated in the Lemma as $A_{l,m,1}$ and $A_{l,m,3}$, for $m > 0$.

Next, assume $m \ge 0$ and consider the case $m' = m + 1$. The integral in \bsrefeqn{eqn:lemmaVSHrecurrencesproof} is then:
\bseqnnumber{
	&\int_{-1}^{1} \Big\{ m(m+1) \jaclm{l}{m} \jaclmbracket{l'}{m+1} \rho^{2m} (1 + z^2) - m z \jaclm{l}{m} \jaclmbracketdz{l'}{m+1} \rho^{2m+2} \nonumber \\
	&\quadtwo \quad - (m+1) z \jaclmdz{l}{m} \jaclmbracket{l'}{m+1} \rho^{2m+2} + \jaclmdz{l}{m} \jaclmbracketdz{l'}{m+1} \rho^{2m+4} \Big\} \: \D z \nonumber \\
	&= \int_{-1}^{1} \Big\{ m(m+1) \jaclm{l}{m} \jaclmbracket{l'}{m+1} \rho^{2m} (1 + z^2) \nonumber \\
	&\quadfour - (m+1) z \jaclmdz{l}{m} \jaclmbracket{l'}{m+1} \rho^{2m+2} \nonumber \\
	&\quadfour + m \jaclmbracket{l'}{m+1} \rho^{2m} \big( - 2(m+1) z^2 \jaclm{l}{m} + \jaclm{l}{m} \rho^{2} - z \jaclmdz{l}{m} \rho^{2} \big) \nonumber \\
	&\quadfour + \jaclmdz{l}{m} \jaclmbracketdz{l'}{m+1} \rho^{2m+2} \Big\} \: \D z \nonumber \\
	&= \int_{-1}^{1} \Big\{ m(m+2) \jaclm{l}{m} \jaclmbracket{l'}{m+1} \rho^{2m+2} - z \jaclmdz{l}{m} \jaclmbracket{l'}{m+1} \rho^{2m+2} \nonumber \\
	&\quadfour + \jaclmdz{l}{m} \jaclmbracketdz{l'}{m+1} \rho^{2m+4} \Big\} \: \D z \nonumber \\
	&= \int_{-1}^{1} \Big\{ m(m+2) \jaclm{l}{m} \jaclmbracket{l'}{m+1} \rho^{2m+2} - d_{l,m} z \jaclmbracket{l}{m+1} \jaclmbracket{l'}{m+1} \rho^{2m+2} \nonumber \\
	&\quadfour + d_{l,m} d_{l',m+1} \jaclmbracket{l}{m+1} \jaclmbracket{l'}{m+2} \rho^{2m+4} \Big\} \: \D z \nonumber \\
	&= \int_{-1}^{1} \Big\{ m(m+2) \jaclmbracket{l'}{m+1} \big( \tilde \alpha_{l, m, 2} \jaclmbracket{l-1}{m+1} + \tilde \alpha_{l, m, 4} \jaclmbracket{l+1}{m+1} \big) \rho^{2m+2} \nonumber \\
	&\quadtwo - d_{l,m} \jaclmbracket{l'}{m+1} \big( \tilde \gamma_{l,m+1,1} \jaclmbracket{l-1}{m+1} + \tilde \gamma_{l,m+1,2} \jaclmbracket{l+1}{m+1} \big) \rho^{2m+2} \nonumber \\
	&\quadtwo + d_{l,m} d_{l',m+1} \jaclmbracket{l}{m+1} \big( \tilde \alpha_{l', m+2, 1} \jaclmbracket{l'-1}{m+1} + \tilde \alpha_{l', m+2, 3} \jaclmbracket{l'+1}{m+1}  \big) \rho^{2m+2} \Big\} \: \D z \nonumber \\
	&= {\delta_{l',l-1} \over 2\pi} \Big\{ \big[ m(m+2) \tilde \alpha_{l, m, 2} - d_{l,m} \tilde \gamma_{l,m+1,1} \big] {1 \over (c_{l-1}^{m+1})^2} + d_{l,m} d_{l-1,m+1} {\tilde \alpha_{l-1, m+2, 3} \over (c_{l}^{m+1})^2} \Big\} \nonumber \\
	&\quadtwo + {\delta_{l',l+1} \over 2\pi} \Big\{ \big[ m(m+2) \tilde \alpha_{l, m, 4} - d_{l,m} \tilde \gamma_{l,m+1,2} \big] {1 \over (c_{l-1}^{m+1})^2} + d_{l,m} d_{l+1,m+1} {\tilde \alpha_{l+1, m+2, 1} \over (c_{l}^{m+1})^2} \Big\}. \label{eqn:lemmaVSHrecurrencesproof3}
}
Again, we can see that the coefficients are then also only possibly nonzero for $l' = l \pm 1$. Combining the above with the \bsrefeqn{eqn:lemmaVSHrecurrencesproof}, we have that our two coefficients for $\calA_{l-1,m-1}, \calA_{l+1,m-1}$ are those stated in the Lemma as $A_{l,m,2}$ and $A_{l,m,4}$, for $m \ge 0$.

@@

The framework we have outlined for the spherical harmonics and the vector spherical harmonics allows us to easily and explicitly derive sparse operator matrices for partial differential operators such as the spherical gradient, divergence and Laplace--Beltrami, as well as other key operators found in examples such as the linearised shallow water equations. These operators will take coefficients of a function's expansion in either the SSH or VSH basis, to coefficients in either basis.

Let $\uvec$ be a vector-valued function on the tangent bundle of $\Omega$ with coefficients vectors $\uvecc, \uveccperp$ for its expansion in the VSH basis up to order $N \in \N$, and let $f$ be a scalar function on $\Omega$ with coefficients vector $\fvecc$ for its expansion in the SSH basis also up to order $N$, i.e.
\bseqn{
	\uvec(x,y,z) &\approx \VSHvecfull(x,y,z)^\top \: \uvec^c \\
	&\equiv \VSHvec(x,y,z)^\top \: \uvecc + \VSHvecperp(x,y,z)^\top \: \uveccperp \\
	&:=  \sum^N_{l=0} \sum^l_{m=-l} \big[u^{\gradY}_{l,m} \, \gradYlm(x,y,z)  + u^{\gradpY}_{l,m} \, \gradpYlm(x,y,z) \big], \\
	f(x,y,z) &\approx \SHvec(x,y,z)^\top \: \fvecc :=  \sum^N_{l=0} \sum^l_{m=-l} f_{l,m} \, \Ylm(x,y,z).
}
In order to fully exploit the sparsity of the relationships and ensure we are using banded-block-banded operators, in problems where we have an interdependency of scalar and vector-valued functions it can be useful to \enquote{expand} the scalar function's coefficients vector $\fvecc$ to be the length of the vector-valued one $\uvec^c$, with zeros being inserted after each entry, i.e. define the \textit{expanded coefficients for a scalar function on $\Omega$ in the SH basis} by
\bseqn{
	\fveccexp :=
		\begin{pmatrix}
			\fveccexp_0 \\
			\vdots \\
			\fveccexp_N
		\end{pmatrix} \in \R^{2(N+1)^2}, \quad
	\fveccexp_l := 
		\begin{pmatrix}
			f_{l,-l} \\
			0 \\
			f_{l,-l+1} \\
			0 \\
			\vdots \\
		 	f_{l,l} \\
			0
		\end{pmatrix} \in \R^{4(l+1)}.
}
To help with the definitions of the operators we will discuss, let $\calE_s, \calE_v$ be (severely) rectangular matrices defined such that $\calE_s \fveccexp = \fvecc$ and $\calE_v \fvecc = \fveccexp$. We are now in a position to derive the sparse differential operators that can be applied to the functions $\uvec, f$.

@@

Let $u : \Omega \to \C$ represent the temperature at a point $(x,y,z)$ on the sphere and suppose that it is modelled by the equation
\bseqnnumber{
	\ppx{t}{u} = \alpha \DeltaS u \label{eqn:heateqn}
}
for some $\alpha \in \R$. A simple timestepping method we can use to iterate with is the BDF2 method \cite[§2.3]{iserles2009first}. Let $u_n$ for $n = 0, 1, 2, \dots$ be the solution at time $n \Delta t$ where $\Delta t$ is the timestep. The are timestepping method is \nomenclature[Delta t]{$\Delta t$}{A timestep for a simulation}
\bseqn{
	{u_1 - u_0 \over \Delta t} &= \alpha \DeltaS u_1 \\
	{3u_{n+1} - 4u_n + u_{n-1} \over 2\Delta t} &= \alpha \DeltaS u_{n+1}, \quad n = 1,2,\dots
}
$\iff$
\bseqn{
	[1 - \Delta t \alpha \DeltaS] u_1 -  &= u_0 \\
	[3 - 2\Delta t \alpha \DeltaS] u_{n+1} &= 4u_n - u_{n-1}, \quad n = 1,2,\dots.
}
Let us write this in coefficient space. Firstly, our coefficients vectors for the function $u$ we will denote $\uvec^c$ -- that is:
\bseqn{
	u(x,y,z) &\approx \SHvec(x,y,z)^\top \: \uvec^c.
}
Our system then becomes
\bseqn{
	[I - \Delta t \alpha \calL] \uvec^c_1 -  &= u_0 \\
	[3I - 2\Delta t \alpha \calL] \uvec^c_{n+1} &= 4\uvec^c_n - \uvec^c_{n-1}, \quad n = 1,2,\dots.
}
From here, it is clear that eat each timestep we simply have a fairly trivial linear system to solve, with the matrices $I - \Delta t \alpha \calL$, $3I - 2\Delta t \alpha \calL$ being banded-block-banded (in fact, simply diagonal here). We demonstrate this theory in \bsreffig{fig:heateqn}, where we see four snapshots of the solution to the heat equation in \bsrefeqn{eqn:heateqn} on the unit sphere with our order $N$ set large enough (here, $N=20$). We set $\alpha = {1 \over 42}$ and a step size of $\Delta t = 0.01$, and take the initial condition for the heat $u$ to be a \enquote{football function}\footnote{These initial conditions are taken from an example found at https://www.chebfun.org/examples/sphere/SphereHeatConduction.html} -- that is we take
\bseqn{
	u_0 := Y_6^0 + \sqrt{14 \over 11} Y_6^6.
}
The snapshots of the solution $u$ in \bsreffig{fig:heateqn} are taken of the initial conditions, plus after $25$, $50$ and $75$ timesteps respectively.

@@

The linearised shallow water equations allow us to showcase more of the differential operators that we defined in \bsrefdef{def:SHoperators} (namely the divergence and gradient operators) and demonstrate how the natural sparsity that our framework brings to the problem leads to a simple sparse linear system to solve.

>>>>

Before we proceed further, let us build up to our definition of the spherical harmonics. We first need to introduce a few classical orthogonal polynomials. Orthogonal polynomials are defined as such by the condition that each polynomial is orthogonal to each other, with respect to a given inner product. In particular, they are orthogonal with respect to all lower degree polynomials.

@@

An important property of the Jacobi matrices is that they are sparse -- specifically they are banded-block-banded matrices.

@@

Three-term recurrence relations for orthogonal polynomials are well established (see for example \cite[\S18.9]{DLMF}), including for multidimensional OPs \cite[\S3.2]{dunkl2014orthogonal}. In a similar vein, we can obtain a recurrence relation for the spherical harmonics by combining each system in \bsrefeqn{eqn:jacobimatsdef} in a specific way. It should be emphasised at this point that this is not necessarily the optimal way of computing the spherical harmonics; various efficient methods and algorithms have been proposed (see e.g. \cite{guy1998efficient, sloan2013efficient, fukushima2012numerical}). We present this method in order to be able to generalise it to other multidimensional OP families later on in this thesis.

@@

For each $l = 0,1,2\dots$ let $\Dlt$ be any matrix that is a left inverse of $A_l$, i.e. such that $\Dlt A_l = \idmat{2l+3}$. Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the $\Dlt$'s, we obtain a lower triangular system \cite[\S3.2]{dunkl2014orthogonal}, which can be expanded to obtain the three-term recurrence, for $l = 0,1,2,\dots$:
\bseqn{
	\begin{cases}
		\SHvec_{-1}(x,y,z) := 0 \\
		\SHvec_{0}(x,y,z) := Y_0 \\
		\SHvec_{l+1}(x,y,z) = -\Dlt (B_l-G_l(x,y,z)) \SHvec_l(x,y,z) - \Dlt C_l  \,\SHvec_{l-1}(x,y,z)
	\end{cases}
}

@@

For the purposes of this discussion, we will only consider the tangent bundle, and thus wish to write down a basis for vector-valued functions with values in the tangent bundle. Fortunately, there is a simple solution known as the vector spherical harmonics (VSHs) \cite{barrera1985vector}: \nomenclature[grad]{$\gradS$}{The spherical gradient operator} \nomenclature[Psilm]{$\gradYlm$}{The vector spherical harmonic $\gradS \Ylm$} \nomenclature[Philm]{$\gradpYlm$}{The vector spherical harmonic $\gradS^\perp \Ylm = \rvec \times \gradS \Ylm$}

@@

On that note, we will first consider the case that $m > 0$ and $m' = m - 1$, and calculate the integral in \bsrefeqn{eqn:lemmaVSHrecurrencesproof}:
\bseqnnumber{
	&\int_{-1}^{1} \Big\{ m(m-1) \jaclm{l}{m} \jaclmbracket{l'}{m-1} \rho^{2m-2} (1 + z^2) - m z \jaclm{l}{m} \jaclmbracketdz{l'}{m-1} \rho^{2m} \nonumber \\
	&\quadtwo \quad - (m-1) z \jaclmdz{l}{m} \jaclmbracket{l'}{m-1} \rho^{2m} + \jaclmdz{l}{m} \jaclmbracketdz{l'}{m-1} \rho^{2m+2} \Big\} \: \D z \nonumber \\
	&= \int_{-1}^{1} \Big\{ m(m-1) \jaclm{l}{m} \jaclmbracket{l'}{m-1} \rho^{2m-2} (1 + z^2) - m z \jaclm{l}{m} \jaclmbracketdz{l'}{m-1} \rho^{2m} \nonumber \\
	&\quadtwo + (m-1) \jaclm{l}{m} \rho^{2m-2} \big(\jaclmbracket{l'}{m-1} \rho^{2} - z \jaclmbracketdz{l'}{m-1} \rho^{2} - 2m z^2 \jaclmbracket{l'}{m-1} \big) \nonumber \\
	&\quadtwo + \jaclmdz{l}{m} \jaclmbracketdz{l'}{m-1} \rho^{2m+2} \Big\} \: \D z \nonumber \\
	&= \int_{-1}^{1} \Big\{ (m-1)(m+1) \jaclm{l}{m} \jaclmbracket{l'}{m-1} \rho^{2m} - z \jaclm{l}{m} \jaclmbracketdz{l'}{m-1} \rho^{2m} \nonumber \\
	&\quadfour + \jaclmdz{l}{m} \jaclmbracketdz{l'}{m-1} \rho^{2m+2} \Big\} \: \D z \nonumber \\
	&= \int_{-1}^{1} \Big\{ (m-1)(m+1) \jaclm{l}{m} \jaclmbracket{l'}{m-1} \rho^{2m} - d_{l',m-1} z \jaclm{l}{m} \jaclm{l'}{m} \rho^{2m} \nonumber \\
	&\quadfour + d_{l,m} d_{l',m-1} \jaclmbracket{l}{m+1} \jaclm{l'}{m} \rho^{2m+2} \Big\} \: \D z \nonumber \\
	&= \int_{-1}^{1} \Big\{ (m^2 - 1) \jaclm{l}{m} \big( \tilde \alpha_{l', m-1, 2} \jaclm{l'-1}{m} + \tilde \alpha_{l', m-1, 4} \jaclm{l'+1}{m} \big) \rho^{2m} \nonumber \\
	&\quadfour - d_{l',m-1} \jaclm{l}{m} \big( \tilde \gamma_{l',m,1} \jaclm{l'-1}{m} + \tilde \gamma_{l',m,2} \jaclm{l'+1}{m} \big) \rho^{2m} \nonumber \\
	&\quadfour + d_{l,m} d_{l',m-1} \jaclm{l'}{m} \big( \tilde \alpha_{l, m+1, 1} \jaclm{l-1}{m} + \tilde \alpha_{l, m+1, 3} \jaclm{l+1}{m} \big) \rho^{2m} \Big\} \: \D z \nonumber \\
	&= {\delta_{l',l-1} \over 2\pi} \Big\{ \big[ (m^2 - 1) \tilde \alpha_{l-1, m-1, 4} - d_{l-1,m-1} \tilde \gamma_{l-1,m,2} \big] {1 \over (\clm)^2} + d_{l,m} d_{l-1,m-1} {\tilde \alpha_{l-1, m-1, 1} \over (c_{l-1}^m)^2} \Big\} \nonumber \\
	&\quadtwo + {\delta_{l',l+1} \over 2\pi} \Big\{ \big[ (m^2 - 1) \tilde \alpha_{l+1, m-1, 2} - d_{l+1,m-1} \tilde \gamma_{l+1,m,1} \big] {1 \over (\clm)^2} \nonumber \\
	&\quadeight + d_{l,m} d_{l+1,m-1} {\tilde \alpha_{l-1, m-1, 3} \over (c_{l+1}^m)^2} \Big\}. \label{eqn:lemmaVSHrecurrencesproof2}
}
As we can see, the coefficients zero for $l' \ne l \pm 1$. Combining the above with the \bsrefeqn{eqn:lemmaVSHrecurrencesproof}, we have that our two coefficients for $\calA_{l-1,m-1}, \calA_{l+1,m-1}$ are those stated in the Lemma as $A_{l,m,1}$ and $A_{l,m,3}$, for $m > 0$.

Next, assume $m \ge 0$ and consider the case $m' = m + 1$. The integral in \bsrefeqn{eqn:lemmaVSHrecurrencesproof} is then:
\bseqnnumber{
	&\int_{-1}^{1} \Big\{ m(m+1) \jaclm{l}{m} \jaclmbracket{l'}{m+1} \rho^{2m} (1 + z^2) - m z \jaclm{l}{m} \jaclmbracketdz{l'}{m+1} \rho^{2m+2} \nonumber \\
	&\quadtwo \quad - (m+1) z \jaclmdz{l}{m} \jaclmbracket{l'}{m+1} \rho^{2m+2} + \jaclmdz{l}{m} \jaclmbracketdz{l'}{m+1} \rho^{2m+4} \Big\} \: \D z \nonumber \\
	&= \int_{-1}^{1} \Big\{ m(m+1) \jaclm{l}{m} \jaclmbracket{l'}{m+1} \rho^{2m} (1 + z^2) \nonumber \\
	&\quadfour - (m+1) z \jaclmdz{l}{m} \jaclmbracket{l'}{m+1} \rho^{2m+2} \nonumber \\
	&\quadfour + m \jaclmbracket{l'}{m+1} \rho^{2m} \big( - 2(m+1) z^2 \jaclm{l}{m} + \jaclm{l}{m} \rho^{2} - z \jaclmdz{l}{m} \rho^{2} \big) \nonumber \\
	&\quadfour + \jaclmdz{l}{m} \jaclmbracketdz{l'}{m+1} \rho^{2m+2} \Big\} \: \D z \nonumber \\
	&= \int_{-1}^{1} \Big\{ m(m+2) \jaclm{l}{m} \jaclmbracket{l'}{m+1} \rho^{2m+2} - z \jaclmdz{l}{m} \jaclmbracket{l'}{m+1} \rho^{2m+2} \nonumber \\
	&\quadfour + \jaclmdz{l}{m} \jaclmbracketdz{l'}{m+1} \rho^{2m+4} \Big\} \: \D z \nonumber \\
	&= \int_{-1}^{1} \Big\{ m(m+2) \jaclm{l}{m} \jaclmbracket{l'}{m+1} \rho^{2m+2} - d_{l,m} z \jaclmbracket{l}{m+1} \jaclmbracket{l'}{m+1} \rho^{2m+2} \nonumber \\
	&\quadfour + d_{l,m} d_{l',m+1} \jaclmbracket{l}{m+1} \jaclmbracket{l'}{m+2} \rho^{2m+4} \Big\} \: \D z \nonumber \\
	&= \int_{-1}^{1} \Big\{ m(m+2) \jaclmbracket{l'}{m+1} \big( \tilde \alpha_{l, m, 2} \jaclmbracket{l-1}{m+1} + \tilde \alpha_{l, m, 4} \jaclmbracket{l+1}{m+1} \big) \rho^{2m+2} \nonumber \\
	&\quadtwo - d_{l,m} \jaclmbracket{l'}{m+1} \big( \tilde \gamma_{l,m+1,1} \jaclmbracket{l-1}{m+1} + \tilde \gamma_{l,m+1,2} \jaclmbracket{l+1}{m+1} \big) \rho^{2m+2} \nonumber \\
	&\quadtwo + d_{l,m} d_{l',m+1} \jaclmbracket{l}{m+1} \big( \tilde \alpha_{l', m+2, 1} \jaclmbracket{l'-1}{m+1} + \tilde \alpha_{l', m+2, 3} \jaclmbracket{l'+1}{m+1}  \big) \rho^{2m+2} \Big\} \: \D z \nonumber \\
	&= {\delta_{l',l-1} \over 2\pi} \Big\{ \big[ m(m+2) \tilde \alpha_{l, m, 2} - d_{l,m} \tilde \gamma_{l,m+1,1} \big] {1 \over (c_{l-1}^{m+1})^2} + d_{l,m} d_{l-1,m+1} {\tilde \alpha_{l-1, m+2, 3} \over (c_{l}^{m+1})^2} \Big\} \nonumber \\
	&\quadtwo + {\delta_{l',l+1} \over 2\pi} \Big\{ \big[ m(m+2) \tilde \alpha_{l, m, 4} - d_{l,m} \tilde \gamma_{l,m+1,2} \big] {1 \over (c_{l-1}^{m+1})^2} + d_{l,m} d_{l+1,m+1} {\tilde \alpha_{l+1, m+2, 1} \over (c_{l}^{m+1})^2} \Big\}. \label{eqn:lemmaVSHrecurrencesproof3}
}
Again, we can see that the coefficients hence zero for $l' \ne l \pm 1$. Combining the above with the \bsrefeqn{eqn:lemmaVSHrecurrencesproof}, we have that our two coefficients for $\calA_{l-1,m-1}, \calA_{l+1,m-1}$ are those stated in the Lemma as $A_{l,m,2}$ and $A_{l,m,4}$, for $m \ge 0$.

@@

The framework we have outlined for the spherical harmonics and the vector spherical harmonics allows us to easily and explicitly derive sparse operator matrices for partial differential operators such as the spherical gradient, divergence and Laplace--Beltrami, as well as other key operators found in examples such as the linearised shallow water equations. These operators will take coefficients of a function's expansion in either the SH or VSH basis, to coefficients in either basis.

Let $\uvec$ be a vector-valued function on the tangent bundle of $\Omega$ with coefficients vectors $\uvecc, \uveccperp$ for its expansion in the VSH basis up to order $N \in \N$, and let $f$ be a scalar function on $\Omega$ with coefficients vector $\fvecc$ for its expansion in the SH basis also up to order $N$, i.e.
\bseqn{
	\uvec(x,y,z) &\approx \VSHvecfull(x,y,z)^\top \: \uvec^c \\
	&\equiv \VSHvec(x,y,z)^\top \: \uvecc + \VSHvecperp(x,y,z)^\top \: \uveccperp \\
	&:=  \sum^N_{l=0} \sum^l_{m=-l} \big[u^{\gradY}_{l,m} \, \gradYlm(x,y,z)  + u^{\gradpY}_{l,m} \, \gradpYlm(x,y,z) \big], \\
	f(x,y,z) &\approx \SHvec(x,y,z)^\top \: \fvecc :=  \sum^N_{l=0} \sum^l_{m=-l} f_{l,m} \, \Ylm(x,y,z).
}
In order to fully exploit the sparsity of the relationships and ensure we are using banded-block-banded operators, in problems where we have an interdependency of scalar and vector-valued functions it can be useful to \enquote{expand} the scalar function's coefficients vector $\fvecc$ to be the length of the vector-valued one $\uvec^c$, with zeros being inserted after each entry, i.e. define the \textit{expanded coefficients for a scalar function on $\Omega$ in the SH basis} by
\bseqn{
	\fveccexp :=
		\begin{pmatrix}
			\fveccexp_0 \\
			\vdots \\
			\fveccexp_N
		\end{pmatrix} \in \R^{2(N+1)^2}, \quad
	\fveccexp_l := 
		\begin{pmatrix}
			f_{l,-l} \\
			0 \\
			f_{l,-l+1} \\
			0 \\
			\vdots \\
		 	f_{l,l} \\
			0
		\end{pmatrix} \in \R^{4(l+1)}.
}
To help with the definitions of the operators we will discuss, let $\calE_s, \calE_v$ be (severely) rectangular matrices defined such that $\calE_s \fveccexp = \fvecc$ and $\calE_v \fvecc = \fveccexp$. We are now in a position to derive the sparse differential operators that can be applied to the functions $\uvec, f$.

@@

Let $u : \Omega \to \C$ represent the temperature at a point $(x,y,z)$ on the sphere and suppose that it is modelled by the equation
\bseqnnumber{
	\ppx{t}{u} = \alpha \DeltaS u \label{eqn:heateqn}
}
for some $\alpha \in \R$. A simple timestepping method we can use to iterate with is the BDF2 method \cite[\S2.3]{iserles2009first}. Let $u_n$ for $n = 0, 1, 2, \dots$ be the solution at time $n \Delta t$ where $\Delta t$ is the timestep. The timestepping method is \nomenclature[Delta t]{$\Delta t$}{A timestep for a simulation}
\bseqn{
	{u_1 - u_0 \over \Delta t} &= \alpha \DeltaS u_1 \\
	{3u_{n+1} - 4u_n + u_{n-1} \over 2\Delta t} &= \alpha \DeltaS u_{n+1}, \quad n = 1,2,\dots
}
$\iff$
\bseqn{
	[1 - \Delta t \alpha \DeltaS] u_1 -  &= u_0 \\
	[3 - 2\Delta t \alpha \DeltaS] u_{n+1} &= 4u_n - u_{n-1}, \quad n = 1,2,\dots.
}
Let us write this in coefficient space. Firstly, our coefficients vectors for the function $u$ we will denote $\uvec^c$ -- that is:
\bseqn{
	u(x,y,z) &\approx \SHvec(x,y,z)^\top \: \uvec^c.
}
Our system then becomes
\bseqn{
	[I - \Delta t \alpha \calL] \uvec^c_1 -  &= u_0 \\
	[3I - 2\Delta t \alpha \calL] \uvec^c_{n+1} &= 4\uvec^c_n - \uvec^c_{n-1}, \quad n = 1,2,\dots.
}
From here, it is clear that at each timestep we simply have a fairly trivial linear system to solve, with the matrices $I - \Delta t \alpha \calL$, $3I - 2\Delta t \alpha \calL$ being banded-block-banded (in fact, simply diagonal here). We demonstrate this theory in \bsreffig{fig:heateqn}, where we see four snapshots of the solution to the heat equation in \bsrefeqn{eqn:heateqn} on the unit sphere with our order $N$ set large enough (here, $N=20$). We set $\alpha = {1 \over 42}$ and a step size of $\Delta t = 0.01$, and take the initial condition for the heat $u$ to be a \enquote{football function}\footnote{These initial conditions are taken from an example found at https://www.chebfun.org/examples/sphere/SphereHeatConduction.html} -- that is we take
\bseqn{
	u_0 := Y_6^0 + \sqrt{14 \over 11} Y_6^6.
}
The snapshots of the solution $u$ in \bsreffig{fig:heateqn} are taken of the initial conditions, plus after $25$, $50$ and $75$ timesteps respectively.

@@

The linearised shallow water equations (SWEs) allow us to showcase more of the differential operators that we defined in \bsrefdef{def:SHoperators} (namely the divergence and gradient operators) and demonstrate how the natural sparsity that our framework brings to the problem leads to a simple sparse linear system to solve.
